---
title: "Assignment 1.2"
author: "Xuening Tang, group5"
date: "18th February 2025"
output: pdf_document
highlight: tango
editor_options: 
  markdown: 
    wrap: 72
---

*In order not to be bothered with rounding the numbers, set
`options(digits=3)r options(digits=3)`.*

## Exercise 2

#### Section a

To study the effect of County and Related on the variable Crops, we
propose the following hypotheses:\
$H_{AB}$ There is no interaction effect of *County and Related*.\
$H_A$ There is no main effect of factor *County*.\
$H_B$ There is no main effect of factor *Related*.

```{r}
library(dplyr)
crop_data <- read.delim("crops.txt", sep = " ") # read the dataset

cropframe = data.frame(
  crops = crop_data$Crops, 
  county = factor(crop_data$County),
  related = factor(crop_data$Related)) # turn variables "county" and "related" into factors; store relevant variables to a new data frame

cropframe <- cropframe %>% mutate(related = recode(related, 'yes' = '1', 'no' = '0')) # recode the "related" variable
```

```{r}
cropframe # display the frame
crops = cropframe$crops
county = cropframe$county
related = cropframe$related
```

Before conducting the ANOVA test, we first plotted two interaction plots
to get a first glimpse of the potential interaction effect. Based on the
two interaction plots, it seems there is little interaction effect, as
the lines are parallel in general. We then conduct a two-way ANOVA to
confirm our observation.

```{r, fig.height=6, fig.width=6}
interaction.plot(related,county,crops) # fix county
interaction.plot(county,related,crops) # fix related
```

```{r}
is.factor(county) # check if county and related are factors
is.factor(related)
```

```{r}
cropanov=lm(crops~county*related); anova(cropanov) # conduct the two-way ANOVA test
summary(cropanov)
```

Result shows that there is no interaction effect between *Related* and
*County* on *Crops*. None of the p values for *County*, *Crops* and
*County:Related* are significant (p = 0.477; p = 0.527; p = 0.879). To
make sure that this result is valid, we plot a Q-Q plot and residual
plot. Based on the two plots, the assumption of normality is met: Q-Q
plot gives a straight line in general, and the residuals display no
pattern.

```{r}
par(mfrow=c(1,2))
qqnorm(residuals(cropanov)); plot(fitted(cropanov),residuals(cropanov))
```

In the next step, we remove the interaction and apply an additive model.
The code and results are shown below:

```{r}
cropanov2=lm(crops~county+related,data=cropframe); anova(cropanov2) # additive model
summary(cropanov2)
```

The result of the additive model shows that neither of the factors has a
significant main effect on Crops. The p-values are 0.4518 and 0.5126 for
*County* and *Related* respectively, and are larger than the 0.05
significance level in both cases. Therefore, we fail to reject none of
our hypotheses. The normality assumption of this ANOVA test is also met
based on the following Q-Q plot and residual plot:

```{r}
qqnorm(residuals(cropanov2)); plot(fitted(cropanov2),residuals(cropanov2))
```

Summary for the decisions to the null hypotheses:

| Hypothesis | Decision   |
|------------|------------|
| $H_{AB}$   | not reject |
| $H_{A}$    | not reject |
| $H_{B}$    | not reject |

The mathematical formula for a two-way ANOVA model is:

$$
Y_{ijk} = \mu_{ij} + e_{ijk}
$$

where

$$
\mu_{ij} = \mu + \alpha_i + \beta_j + \gamma_{ij}
$$

$\mu$ is the overall mean

$\alpha_i$ is the main effect of level i of the factor *County*, i =
1,2,3

$\beta_j$ is the main effect of level j of the factor *Related*, j = 0,1

$\gamma_ij$ is the interaction effect of levels i, j of factor *County*
and *Related*, which is 0 in this case, since there is no significant
interaction effect.

We apply the model **cropanov2** for prediction.Therefore, the crops in
*County 3* for which there is no related is:

$$
Crops = Intercept + County3 + Related0 = 6800.6 + 959.7 + 0 = 7760.3
$$

Therefore, the predicted value of the Crops is 7760.3.

#### Section b

We now add the variable *Size*. Since it is a numerical variable, it
should be treated as an exploratory variable. We want to find out how
*Size* influences the effect of *Related* or *County* on *Crops* in our
model.

(1) ANCOVA test: *Size \* County*

$H_{AB}$ There is no interaction effect between *Size* and *County* on
*Crops*

We first get a glimpse of the distribution of size in different
counties. It seems the distributions are different in different
counties. We need to confirm our observation through a two-way ANCOVA
model.

```{r}
size = crop_data$Size
boxplot(size~county)
```

```{r}
cropanov3=lm(crops~size*county,data=cropframe);anova(cropanov3)
summary(cropanov3)
```

Based on the result, there is a significant interaction effect between
*Size* and *County* on *Crops (p-value = 0.007)*. Summary of the ANOVA
model shows that the effect mainly lies on the combination of
*size:county 2 (p-value = 0.002)*, while size:county 3 is not
significant (p-value = 0.157). Meanwhile, *county 2* also has a
significant main effect under the influence of *Size* (p-value = 0.005).
Q-Q plot and residual plot show that the assumption of normality is met
in this case:

```{r}
qqnorm(residuals(cropanov3)); plot(fitted(cropanov3),residuals(cropanov3))
```

We **reject the null hypothesis** that there is no interaction effect
between *Size* and *County*

2.  ANCOVA test: *Size \* Related*

$H_{AB}$ There is no interaction effect between *Size* and *Related* on
*Crops*

```{r}
boxplot(size~related)
```

```{r}
cropanov4=lm(crops~size*related,data=cropframe);anova(cropanov4)
summary(cropanov4)
```

In the box plot, the distribution of *Size* does not differ much for
different *Related* values. The interaction effect is also not
significant according to the result of the ANCOVA test (p=0.331).
Therefore, we **cannot reject the null hypothesis** in this case. The
assumption of normality is met in general, though one can argue that
there are more residuals on the lower fitted value side.

```{r}
qqnorm(residuals(cropanov4)); plot(fitted(cropanov4),residuals(cropanov4))
```

We then conduct two ANCOVA tests without interaction. We investigate the
main effect of *Related* under the influence of *Size (cropanov5)*, and
the main effect of *Size* under the influence of *Related (cropanov6).*

```{r}
cropanov5=lm(crops~size+related,data=cropframe);anova(cropanov5)# related on the second place
summary(cropanov5)
```

```{r}
cropanov6=lm(crops~related+size,data=cropframe);anova(cropanov6)# size on the second place
summary(cropanov6)
```

Results show that *Related* does not have a significant main effect on
*Crops* under the influence of *Size*, but *Size* has a significant main
effect under the influence of *Related*. The normality assumption of
both of the ANCOVA tests are met in general, although there seems to be
more residuals in the area with a lower fitted score:

```{r}
qqnorm(residuals(cropanov5)); plot(fitted(cropanov5),residuals(cropanov5))
```

```{r}
qqnorm(residuals(cropanov6)); plot(fitted(cropanov4),residuals(cropanov6))
```

Summary of this part:

(1) There is a significant interaction effect between *Size* and
    *County* on *Crops*. The combination *Size:County 2* is making the
    most contribution.

(2) No significant interaction effect is found for *Size \* Related*.

(3) *Related* has no significant main effect on *Crops* under the
    influence of *Size*, while *Size* has a significant main effect on
    *Crops* under the influence of *Related*.

#### Section c

Based on our findings in part (b), we now include all factors (*Related*
and *County*) and the exploratory variable (*Size*) together in the same
model. We conduct a full ANCOVA test.

$$
Crops∼County+Related+Size+(County×Size)+(Related×Size)+(County×Related)
$$

We already know there is no interaction between *Related and Size*, and
*County and Related*, so we drop the last two terms and get:

$$
Crops∼County+Related+Size+(County×Size)
$$

```{r}
cropanov7=lm(crops ~ county+related+size+county*size, data = crop_data);anova(cropanov7)# size on the second place
summary(cropanov7)
```

```{r}
qqnorm(residuals(cropanov7)); plot(fitted(cropanov7),residuals(cropanov7))
```

Result of a full ANCOVA test also confirms what we found in section B in
general. There is a significant main effect of variable *Size (p =
0.000)* and *County 2 (p = 0.008)*. The interaction effect of
*Size:County* is therefore also significant *(p = 0.012)*. Different
from section B, *County* now also has a slightly significant main effect
*(p = 0.016)*, when *County, Related* and *Size* are all included into
the model. The assumption of normality is met according to the Q-Q plot
and residual plot.

Based on the summary, we can derive several conclusions:

-   There is a significant difference between crop values in these three
    counties. County 2 yields significantly fewer crops than County 1
    (Estimate = -4214.050).

-   Size has a significantly positive effect on crops, with a larger
    land results in more crops (Estimate = 22.704).

-   The positive effect of size is more prominent in County 2, as there
    is a significant interaction effect. They yield a higher crops value
    than Size: County1 (Estimate = 26.590).

-   Value of crops does not depend on the relation of landlord and
    tenant in all three counties, ,as the difference between different
    relations is not statistically significant.

The following table summarizes all ANOVA tests we've run in this
exercise: (the ones with bold fonts are used for predictions)

| Name | Test Type | R-model |
|----|----|----|
| cropanov | Two-way ANOVA model | \$ Crops∼(County×Related) \$ |
| **cropanov2** | Additive model | \$ Crops∼(County+Related) \$ |
| cropanov3 | Two-way ANCOVA model | \$ Crops∼(County×Size) \$ |
| cropanov4 | Two-way ANCOVA model | \$ Crops∼(Size×Related) \$ |
| cropanov5 | Additive model | \$ Crops∼(Size+Related) \$ |
| cropanov6 | Additive model | \$ Crops∼(Related+Size) \$ |
| **cropanov7** | Full ANCOVA model | $Crops∼County+Related+Size+(County×Size)$ |

#### Section d

We will apply model **cropanov7** to make the prediction. The
mathematical formula for a full ANCOVA is:

$$ Y_{ijk} = \mu_{ij} + e_{ijk} $$

where

$$
\mu_{ijk} = \mu + \alpha_i + \beta_j + \delta_k + \gamma_{ik}
$$

$\mu$ is the overall mean

$\alpha_i$ is the main effect of level i of the factor *County*, i =
1,2,3

$\beta_j$ is the main effect of level j of the factor *Related*, j = 0,1

$\delta_k$ is the main effect of the exploratory variable Size, k =
1...n

$\gamma_ik$ is the interaction effect of levels i, k of factor *County*
and exploratory variable *Size*.

According to this equation, the crops from County 2 of size 165, and
related landlord and tenant is therefore:

$$
Crops = Intercept + County2 + Related1 + Size 165 + County2*Size165
$$

$$
= 2461.014-4214.050-239.099+22.704*165+26.590*165 = 6141.378
$$

So the final crops value is 6141.378

The error variance is given by:

$$
\hat{\sigma}^2 = \frac{\text{RSS}}{\text{df}}
$$

According to the summary of the **cropanov7,** we then have:$$
\hat{\sigma}^2 = \frac{20277325}{23} = 881623
$$

The error variance is therefore 881623
