---
title: "Experimental Design and Data Analysis - Assignment 1"
author: "Group 5 - Ivana Malčić, Xuening Tang, Xiaoxuan Zhang"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex
---

*In order not to be bothered with rounding the numbers, set `options(digits=3)r options(digits=3)`.*

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Matrix)
```

## Exercise 2

#### Section a

To study the effect of County and Related on the variable Crops, we propose the following hypotheses:\
$H_{AB}$ There is no interaction effect of *County and Related*.\
$H_A$ There is no main effect of factor *County*.\
$H_B$ There is no main effect of factor *Related*.

```{r}
library(dplyr)
crop_data <- read.delim("crops.txt", sep = " ") # read the dataset

cropframe = data.frame(
  crops = crop_data$Crops, 
  county = factor(crop_data$County),
  related = factor(crop_data$Related)) # turn variables "county" and "related" into factors; store relevant variables to a new data frame

cropframe <- cropframe %>% mutate(related = recode(related, 'yes' = '1', 'no' = '0')) # recode the "related" variable
```

```{r}
cropframe # display the frame
crops = cropframe$crops
county = cropframe$county
related = cropframe$related
```

Before conducting the ANOVA test, we first plotted two interaction plots to get a first glimpse of the potential interaction effect. Based on the two interaction plots, it seems there is little interaction effect, as the lines are parallel in general. We then conduct a two-way ANOVA to confirm our observation.

```{r, fig.height=6, fig.width=6}
interaction.plot(related,county,crops) # fix county
interaction.plot(county,related,crops) # fix related
```

```{r}
is.factor(county) # check if county and related are factors
is.factor(related)
```

```{r}
cropanov=lm(crops~county*related); anova(cropanov) # conduct the two-way ANOVA test
summary(cropanov)
```

Result shows that there is no interaction effect between *Related* and *County* on *Crops*. None of the p values for *County*, *Crops* and *County:Related* are significant (p = 0.477; p = 0.527; p = 0.879). To make sure that this result is valid, we plot a Q-Q plot and residual plot. Based on the two plots, the assumption of normality is met: Q-Q plot gives a straight line in general, and the residuals display no pattern.

```{r}
par(mfrow=c(1,2))
qqnorm(residuals(cropanov)); plot(fitted(cropanov),residuals(cropanov))
```

In the next step, we remove the interaction and apply an additive model. The code and results are shown below:

```{r}
cropanov2=lm(crops~county+related,data=cropframe); anova(cropanov2) # additive model
summary(cropanov2)
```

The result of the additive model shows that neither of the factors has a significant main effect on Crops. The p-values are 0.4518 and 0.5126 for *County* and *Related* respectively, and are larger than the 0.05 significance level in both cases. Therefore, we fail to reject none of our hypotheses. The normality assumption of this ANOVA test is also met based on the following Q-Q plot and residual plot:

```{r}
qqnorm(residuals(cropanov2)); plot(fitted(cropanov2),residuals(cropanov2))
```

Summary for the decisions to the null hypotheses:

| Hypothesis | Decision   |
|------------|------------|
| $H_{AB}$   | not reject |
| $H_{A}$    | not reject |
| $H_{B}$    | not reject |

The mathematical formula for a two-way ANOVA model is:

$$
Y_{ijk} = \mu_{ij} + e_{ijk}
$$

where

$$
\mu_{ij} = \mu + \alpha_i + \beta_j + \gamma_{ij}
$$

$\mu$ is the overall mean

$\alpha_i$ is the main effect of level i of the factor *County*, i = 1,2,3

$\beta_j$ is the main effect of level j of the factor *Related*, j = 0,1

$\gamma_ij$ is the interaction effect of levels i, j of factor *County* and *Related*, which is 0 in this case, since there is no significant interaction effect.

We apply the model **cropanov2** for prediction.Therefore, the crops in *County 3* for which there is no related is:

$$
Crops = Intercept + County3 + Related0 = 6800.6 + 959.7 + 0 = 7760.3
$$

Therefore, the predicted value of the Crops is 7760.3.

#### Section b

We now add the variable *Size*. Since it is a numerical variable, it should be treated as an exploratory variable. We want to find out how *Size* influences the effect of *Related* or *County* on *Crops* in our model.

(1) ANCOVA test: *Size \* County*

$H_{AB}$ There is no interaction effect between *Size* and *County* on *Crops*

We first get a glimpse of the distribution of size in different counties. It seems the distributions are different in different counties. We need to confirm our observation through a two-way ANCOVA model.

```{r}
size = crop_data$Size
boxplot(size~county)
```

```{r}
cropanov3=lm(crops~size*county,data=cropframe);anova(cropanov3)
summary(cropanov3)
```

Based on the result, there is a significant interaction effect between *Size* and *County* on *Crops (p-value = 0.007)*. Summary of the ANOVA model shows that the effect mainly lies on the combination of *size:county 2 (p-value = 0.002)*, while size:county 3 is not significant (p-value = 0.157). Meanwhile, *county 2* also has a significant main effect under the influence of *Size* (p-value = 0.005). Q-Q plot and residual plot show that the assumption of normality is met in this case:

```{r}
qqnorm(residuals(cropanov3)); plot(fitted(cropanov3),residuals(cropanov3))
```

We **reject the null hypothesis** that there is no interaction effect between *Size* and *County*

2.  ANCOVA test: *Size \* Related*

$H_{AB}$ There is no interaction effect between *Size* and *Related* on *Crops*

```{r}
boxplot(size~related)
```

```{r}
cropanov4=lm(crops~size*related,data=cropframe);anova(cropanov4)
summary(cropanov4)
```

In the box plot, the distribution of *Size* does not differ much for different *Related* values. The interaction effect is also not significant according to the result of the ANCOVA test (p=0.331). Therefore, we **cannot reject the null hypothesis** in this case. The assumption of normality is met in general, though one can argue that there are more residuals on the lower fitted value side.

```{r}
qqnorm(residuals(cropanov4)); plot(fitted(cropanov4),residuals(cropanov4))
```

We then conduct two ANCOVA tests without interaction. We investigate the main effect of *Related* under the influence of *Size (cropanov5)*, and the main effect of *Size* under the influence of *Related (cropanov6).*

```{r}
cropanov5=lm(crops~size+related,data=cropframe);anova(cropanov5)# related on the second place
summary(cropanov5)
```

```{r}
cropanov6=lm(crops~related+size,data=cropframe);anova(cropanov6)# size on the second place
summary(cropanov6)
```

Results show that *Related* does not have a significant main effect on *Crops* under the influence of *Size*, but *Size* has a significant main effect under the influence of *Related*. The normality assumption of both of the ANCOVA tests are met in general, although there seems to be more residuals in the area with a lower fitted score:

```{r}
qqnorm(residuals(cropanov5)); plot(fitted(cropanov5),residuals(cropanov5))
```

```{r}
qqnorm(residuals(cropanov6)); plot(fitted(cropanov4),residuals(cropanov6))
```

Summary of this part:

(1) There is a significant interaction effect between *Size* and *County* on *Crops*. The combination *Size:County 2* is making the most contribution.

(2) No significant interaction effect is found for *Size \* Related*.

(3) *Related* has no significant main effect on *Crops* under the influence of *Size*, while *Size* has a significant main effect on *Crops* under the influence of *Related*.

#### Section c

Based on our findings in part (b), we now include all factors (*Related* and *County*) and the exploratory variable (*Size*) together in the same model. We conduct a full ANCOVA test.

$$
Crops∼County+Related+Size+(County×Size)+(Related×Size)+(County×Related)
$$

We already know there is no interaction between *Related and Size*, and *County and Related*, so we drop the last two terms and get:

$$
Crops∼County+Related+Size+(County×Size)
$$

```{r}
cropanov7=lm(crops ~ county+related+size+county*size, data = crop_data);anova(cropanov7)# size on the second place
summary(cropanov7)
```

```{r}
qqnorm(residuals(cropanov7)); plot(fitted(cropanov7),residuals(cropanov7))
```

Result of a full ANCOVA test also confirms what we found in section B in general. There is a significant main effect of variable *Size (p = 0.000)* and *County 2 (p = 0.008)*. The interaction effect of *Size:County* is therefore also significant *(p = 0.012)*. Different from section B, *County* now also has a slightly significant main effect *(p = 0.016)*, when *County, Related* and *Size* are all included into the model. The assumption of normality is met according to the Q-Q plot and residual plot.

Based on the summary, we can derive several conclusions:

-   There is a significant difference between crop values in these three counties. County 2 yields significantly fewer crops than County 1 (Estimate = -4214.050).

-   Size has a significantly positive effect on crops, with a larger land results in more crops (Estimate = 22.704).

-   The positive effect of size is more prominent in County 2, as there is a significant interaction effect. They yield a higher crops value than Size: County1 (Estimate = 26.590).

-   Value of crops does not depend on the relation of landlord and tenant in all three counties, ,as the difference between different relations is not statistically significant.

The following table summarizes all ANOVA tests we've run in this exercise: (the ones with bold fonts are used for predictions)

| Name | Test Type | R-model |
|----|----|----|
| cropanov | Two-way ANOVA model | \$ Crops∼(County×Related) \$ |
| **cropanov2** | Additive model | \$ Crops∼(County+Related) \$ |
| cropanov3 | Two-way ANCOVA model | \$ Crops∼(County×Size) \$ |
| cropanov4 | Two-way ANCOVA model | \$ Crops∼(Size×Related) \$ |
| cropanov5 | Additive model | \$ Crops∼(Size+Related) \$ |
| cropanov6 | Additive model | \$ Crops∼(Related+Size) \$ |
| **cropanov7** | Full ANCOVA model | $Crops∼County+Related+Size+(County×Size)$ |

#### Section d

We will apply model **cropanov7** to make the prediction. The mathematical formula for a full ANCOVA is:

$$ Y_{ijk} = \mu_{ij} + e_{ijk} $$

where

$$
\mu_{ijk} = \mu + \alpha_i + \beta_j + \delta_k + \gamma_{ik}
$$

$\mu$ is the overall mean

$\alpha_i$ is the main effect of level i of the factor *County*, i = 1,2,3

$\beta_j$ is the main effect of level j of the factor *Related*, j = 0,1

$\delta_k$ is the main effect of the exploratory variable Size, k = 1...n

$\gamma_ik$ is the interaction effect of levels i, k of factor *County* and exploratory variable *Size*.

According to this equation, the crops from County 2 of size 165, and related landlord and tenant is therefore:

$$
Crops = Intercept + County2 + Related1 + Size 165 + County2*Size165
$$

$$
= 2461.014-4214.050-239.099+22.704*165+26.590*165 = 6141.378
$$

So the final crops value is 6141.378

The error variance is given by:

$$
\hat{\sigma}^2 = \frac{\text{RSS}}{\text{df}}
$$

According to the summary of the **cropanov7,** we then have:$$
\hat{\sigma}^2 = \frac{20277325}{23} = 881623
$$

The error variance is therefore 881623

## Exercise 3: Yield of peas

### Section a

```{r}

library('MASS')

set.seed(123)  # add random seed for reproduce

# initial params
I <- 6  # blocks
J <- 4  # plots per block

# initial data frame
randomized_design <- data.frame(block = rep(1:I, each = J), plot = rep(1:J, times = I))

# for each block b, put (N, P, K) on each 2 plots randomly
for (b in 1:I) {
  plots <- sample(1:J, J, replace = FALSE)  # randomly reorder plots in each block
  
  # put N in the header 2 plots
  randomized_design$N[randomized_design$block == b] <- ifelse(plots %in% plots[1:2], 1, 0)
  
  # randomly put P in 2 plots
  randomized_design$P[randomized_design$block == b] <- ifelse(plots %in% sample(plots, 2), 1, 0)
  
  # randomly put K in 2 plots
  randomized_design$K[randomized_design$block == b] <- ifelse(plots %in% sample(plots, 2), 1, 0)
}

# print the plots
print(randomized_design)


```

### Section b

```{r, fig.height = 6}

# combine those yield in the same block same N, and calc its mean
yield_matrix <- tapply(npk$yield, list(npk$block, npk$N), mean)

# plot bars
barplot(t(yield_matrix), beside = TRUE, col = c("red", "blue"),
        main = "Average Yield per Block for Nitrogen Treatment",
        xlab = "Block", ylab = "Average Yield")
legend("top", legend = c("N=0", "N=1"), fill = c("red", "blue"))


```

This plot illustrates that the average yields for soil treated by N are higher than for untreated soil. What's more, each block and treatment tend to have a similar change.

Meanwhile, we here have assigned treatments randomly to each soil within a block, which reduces the variation and get more precise results.

### Section c

```{r}

data0 = npk
data0$block = as.factor(data0$block)
data0$N = as.factor(data0$N)

```

```{r}

# Two-Way ANOVA
model2way = lm(yield~N*block, data=data0)
anova(model2way)

```

p \> 0.05, which means there is no significant evidence of interaction effect.

```{r, fig.height = 3.5}
  
interaction.plot(data0$N,data0$block,data0$yield)
interaction.plot(data0$block,data0$N,data0$yield)

```

Interaction plot also display parallel lines, indicating no interaction.

So, we have to try the "additive" model:

```{r}

modeladd <- lm(yield ~ N + block, data = data0)
anova(modeladd)

```

In both cases p \< 0.05, so both factors have a main effect in the "additive" model.

```{r, fig.height = 3.5}

# Diagnostics:
par(mfrow=c(1,2)) 
qqnorm(residuals(modeladd)); plot(fitted(modeladd),residuals(modeladd))

```

From QQPlot, we can tell that the curve more or less straight, so it is likely normal. Meanwhile, there is no significant pattern in the fitted plot, which is good and means the residual is independent and identical.

-   **Was it sensible to include factor block into this model?** From the results showed in "additive" model, the p_value of block is 0.007095 \< 0.05, and the N is the first order in our model, so it makes sense to include the block.
-   **Can we also apply the Friedman test for this situation?** No, because each block has more than one same value N, meanwhile, the treatments are not completely randomized.

### Section d

```{r}

pairwiseP <- lm(yield ~ block*P + N + K, data = data0) 
pairwiseK <- lm(yield ~ block*K + P + N, data = data0) 
pairwiseN <- lm(yield ~ block*N + K + P, data = data0)

```

```{r}

anova(pairwiseP); anova(pairwiseK); anova(pairwiseN)

```

No interaction effect for either of the three, thus we do an additive model:

```{r}

modeladd2 <- lm(yield ~ block + N + P + K, data = data0); anova(modeladd2)

```

p \< 0.05 for N and K, showing a main effect. p \> 0.05 for P, so we can conclude that there is no significant effect.

**We conclude that the best model is the additive ANOVA model.** Additive ANOVA model provides an overall indication of the effects of each factor N, P, K and block. From the p values we can tell additive model are better than the pairwise model. Furthermore, the pairwise models only focus on the interaction between two factors at a time, lacking the control for others.

### Section e

```{r}
model_d = lm(yield ~ block + N + K + P, data = data0)
summary(model_d)
```

From the summary we can see block 3 is the best in all blocks, and N1 is better than N0, which means N treated is better, while P and K are prefered to be untreated. So, the best combination is **(3, 1, 0, 0)** for (block, N, P, K), leading the largest yield.

### Section f

```{r}

library('MASS')
library('lme4')
data0 = npk

model_mixed <- lmer(yield ~ N+P+K+(1|block),data=data0,REML=FALSE) 
model_fixed <- lm(yield ~  N + P + K + block, data = data0)
anova(model_mixed,model_fixed) 
```

The model comparison results are:

-   AIC: The fixed effects model (AIC = 143.39) is lower than the mixed effects model (AIC = 151.03), suggesting better model fit.
-   BIC: The fixed effects model (BIC = 155.17) is also lower, reinforcing the AIC results.
-   Log-likelihood: The fixed effects model has a higher log-likelihood (-61.695 vs. -69.514), meaning it fits the data better.
-   Chi-square test: χ² = 15.639, p = 0.003544 (significant at p \< 0.05), indicating that treating block as a fixed effect is more appropriate.

The fixed effects model (lm(yield \~ block + N + P + K)) provides a better fit than the mixed effects model.
