---
title: "Assignment 2.1"
author: "Xuening Tang, group5"
date: "4th March 2025"
output: pdf_document
highlight: tango
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 1

#### Section a

We first give a summary of the data:

```{r}
titanic <- read.delim("titanic.txt", header=TRUE)
age <- as.numeric(titanic$Age)
sex = as.factor(titanic$Sex)
class = as.factor(titanic$PClass)
survive = as.factor(titanic$Survived)
```

```{r}
par(mfrow=c(1,3))
hist(age, col = "green")  # distribution of age

tab <- xtabs(~survive + class, data = titanic)
prop_tab <- prop.table(tab, margin = 1)
barplot(prop_tab, main="survival vs. class",
        col = c("red", "blue"),
        beside = TRUE,
        xlab = "Class", ylab = "Survival") # class vs survival

tab <- xtabs(~survive + sex, data = titanic)
prop_tab1 <- prop.table(tab, margin = 1)
barplot(prop_tab1, main="survival vs.sex",
        col = c("red", "blue"),
        beside = TRUE,
        xlab = "Sex", ylab = "Survival") # Sex vs survival
```

The first bar plot illustrates the distribution of age. Most people were
around the age of 20 to 40. The second bar plot compares the survival
rate in different classes (red: death; blue: survive). Most people in
the 1st class survived, while most people in the 3rd class died. The
plot on the right side compares the survival rate in different sexes
(red: death; blue: survive). A bigger proportion of survived customers
were females, while males contributed to a higher proportion of deaths.

We fitted the data with a logistic regression model, with survival as
the outcome variable, age as the exploratory variable, and class and sex
as the factors. The null hypotheses are:

$H_a$ *Factor class and sex have no influence on the survival status.*

$H_b$ *Explanatory variable age has no influence on the survival
status.*

```{r}
tiglm=glm(survive~class+age+sex,data=titanic,family=binomial)
summary(tiglm)
```

Based on the result of the test, all three variables have **a
significant effect** on the survival status, so both null hypotheses are
rejected. The calculation of odd in this case is:

$$
o_{ik} = \frac{P(Y_{ik} = 1)}{P(Y_{ik} = 0)} = e^{\mu + \alpha_i + \beta X_{ik}} \cong  exp(3.76 -1.29class2 -2.52class3-0.04age-2.63male)
$$

An increase in age by 1 year would increase the odd by $e^{-0.04}$, a
change from female to male will change the odd by $e^{-2.63}$, and a
change from class 1 to class 2 will change the odd by $e^{-1.29}$. Since
the exponential function is monotonous, all those changes will result in
a lower probability of survival.

#### Section b

We investigated the interaction effect between age and class and sex and
age on the survival status. We have the following null hypotheses:

$H_0(1)$ *There is no interaction effect between age and class*

$H_0(2)$ *There is no interaction effect between age and sex*

The results are displayed below:

```{r}
tiglm2=glm(survive~class*age,data=titanic,family=binomial)
summary(tiglm2)
```

```{r}
tiglm3=glm(survive~sex*age,data=titanic,family=binomial)
summary(tiglm3)
```

The result for *age \* class* **is not significant** (*p = 0.405; p =
0.771*). while that for *age \* sex* **is significant** (*p \< 0.001*).
Therefore, we reject $H_0(2)$ but preserve $H_0(1)$. We incorporated the
interaction effect into the model we have in section a:

```{r}
tiglm4=glm(survive~sex*age+class,data=titanic,family=binomial)
summary(tiglm4) 
```

Based on the result of this model, the main effects of sex and age are
**no longer significant** when the interaction effect of *age \* sex* is
incorporated. We predicted the probability of survival for each
combination of levels of the factors class and sex for a person aged 55:

```{r}
new_data = expand.grid(
  age = 55,
  class = c("1st","2nd", "3rd"),
  sex = c("female","male")
)
rownames(new_data) <- c("1st&female","2nd&female","3rd&female",
                        "1st&male","2nd&male","3rd&male")
predict(tiglm4,new_data,type = "response")
```

The probability of survival for a female in the 1st class is the highest
(0.947), while that for a male in the 3rd class is the lowest (0.012).

#### Section c

The formula for the logistic regression is:

$$
P(Y_k = 1) = \frac{1}{1+e^{-x^{T}_{k}\theta}}, k = 1,2…,N
$$

where we want to estimate the parameter $\theta$. We can do this by
splitting our data set into two parts (90% and 10%), where the bigger
part is used for training the model and estimating $\theta$, while the
smaller one is used for testing and validating our estimation. We can
evaluate our model by comparing the result of our prediction with the
true survival status and adjust our estimation based on the performance
of our model.

#### Section d

Another approach to study the main effect of class and sex on the
survival status is conducting Chi-square tests on contingency tables:

```{r}
# create a tab for variable sex and class
tabd <- xtabs(~class+survive, data = titanic)
tabd2 <- xtabs(~sex+survive, data = titanic)
```

```{r}
z = chisq.test(tabd);z
z2 = chisq.test((tabd2));z2
```

According to the Chi-square tests, both the effect of sex and class
**are significant**(*p \<0.001*). However, this result deviates partly
from what we got in section b, where we spotted an insignificant main
effect of sex. This is because we didn't introduce the interaction
effect in this case, unlike in section b. Inclusion of interaction and
other additive variables may change the p-value.

#### Section e

Even though the results we got in sections d and b are not completely
consistent, the approach in section d is not necessarily wrong. A
contingency table can help us test whether two variables are independent
or homogeneous. An advantage is that, before conducting any statistical
test or making any calculation, we may already have a first intuition of
the potential relationship between the two variables. This can be done
by simply comparing values in different cells. The disadvantage is that
we need at least 5 items in each cell to make sure that the Chi-square
result is valid. This can be a problem when we are working with data
that do not have enough items for each category. Sometimes dividing the
data into different categories may not be trivial either. A contingency
table is also harder to interpret when we have multiple variables and
want to study the interaction effects among them.

Logistic regression can incorporate a wide range of variables: binary,
explanatory, and categorical (factors). A drawback is that it assumes a
linear relationship between the outcome and predictors, so it may not
work well for non-linear patterns.
